---
title: "Jiahui Li_Computer assignment 1"
output: pdf_document
date: "2024-09-24"
---
# 1.Task1 - Function
```{r}
#Likelihood function
L <- function(theta,y,X) {
  p <- 1/(1 + exp(-X %*% theta))  # logistic function
  likelihood <- prod(dbinom(y, size = 1, prob = p, log = FALSE)) # likelihood 
  return(likelihood)
}

#Log-likelihood function
l <- function(theta,y,X){
  p <- 1/(1+exp(-X %*% theta))
  log_likelihood <- sum((dbinom(y,size=1,prob=p,log= TRUE)))
  return(log_likelihood)
}

#Score function
S <- function(theta,y,X){
  p <- 1/(1+exp(-X %*% theta))
  score <- t(X) %*% (y-p)
  return(score)
}

#Fisher Information I function
I <- function(theta,y,X){
  p <- 1/(1+exp(-X %*% theta))
  v <- p * (1-p)
  D = diag(as.vector(v))
  fisher_info = t(X) %*% D %*% X
  return(fisher_info)
}

#Newton-Raphson
NR <- function(theta0, niter, y, X, tol = 0.001) {
  theta <- theta0  # Initialize parameters
  theta_history <- matrix(NA, niter + 1, length(theta0))  # Store parameter values for each iteration
  theta_history[1, ] <- theta0  # Store the initial value in history
  
  for (i in 1:niter) {
    score <- S(theta, y, X)  # Compute the score function
    fisher_info <- I(theta, y, X)  # Compute the Fisher Information matrix
    theta_new <- theta + solve(fisher_info) %*% score  # Update the parameters
    
    # Store the new parameter estimates
    theta_history[i + 1, ] <- theta_new
    
    # Compare the difference between new and old parameter values
    if (all(abs(theta_new - theta) < tol)) {
      cat("Converged after", i, "iterations\n")
      return(list(final_theta = theta_new, theta_history = theta_history[1:(i + 1), ]))
    }
    
    # Update theta with the new estimates and continue iterating
    theta <- theta_new
  }
  
  cat("Did not fully converge after", niter, "iterations\n")
  return(list(final_theta = theta, theta_history = theta_history))
}
```

# Data preparation

```{r}
library(RCurl)
set.seed(0106016) #Change 990101 to your date of birth
länk <- "https://raw.githubusercontent.com/mskoldSU/MT5003_HT17/master/Projekt/proj_data.csv"
data_individ <- read.csv(text = getURL(länk))
idx <- sample(1:nrow(data_individ), 1000)
data_individ <- data_individ[idx, ]
save(data_individ, file = "proj_data.Rdata")
head(data_individ)
load("proj_data.Rdata")
modell <- glm(Resultat ~ Alder + Kon + Utbildare, 
              data = data_individ,
              family = "binomial")
summary(modell)
y <- matrix(data_individ$Resultat, ncol = 1)
X <- model.matrix(Resultat ~ Alder + Kon + Utbildare, 
                  data = data_individ)
head(X)
head(data_individ[, -1])
```
## data glm_iterations

```{r}
# A model with all explanatory variables is fitted in R
modell <- glm(Resultat ~ Alder + Kon + Utbildare, 
              data = data_individ,
              family = "binomial")

summary(modell)  # Print R's estimated results

# Calculate the difference between NR estimates and glm estimates after each iteration to find convergence
glm_estimates <- coef(modell)

# Print the number of iterations used by the glm model
iterations <- modell$iter
cat("GLM iterations: ", iterations, "\n")

```
- The model required 4 iterations to converge.
- Age (Alder) has a negative impact on the outcome, while being male (KonMan) and attending a driving school (UtbildareTrafikskola) significantly increase the probability of the positive outcome.
- All the predictors are statistically significant except for the intercept, indicating they play an important role in explaining the variation in the outcome.

# 2.Task2
- Verify your function NR from task 1 by re-computing the parameter estimates provided by R (R actually uses the same type of algorithm). If you supply starting value theta0 = c(0, 0, 0, 0), how many iterations (niter) is needed in order to re-compute Rs estimates with two digits of accuracy?
```{r}

# Define initial values for theta0 and maximum number of iterations
theta0 <- c(0, 0, 0, 0)  # Initial parameter values
niter <- 50  # Maximum number of iterations

# Run the NR algorithm and check if it converges to two decimal places accuracy
result <- NR(theta0, niter, y, X)

# Print the final estimated values
print(result$final_theta)

# Print the history of parameter estimates for each iteration
print(result$theta_history)

# Compare the parameter estimates from each iteration and observe the changes
for (i in 1:(nrow(result$theta_history) - 1)) {
  cat("Iteration", i, "->", i + 1, "Difference:\n")
  print(abs(result$theta_history[i + 1, ] - result$theta_history[i, ]))
}

```

```{r}
parameter_names <- c(
  "Intercept (Parameter 1)", 
  "Alder (Age) (Parameter 2)", 
  "Kon (Gender) (Parameter 3)", 
  "Utbildare (Education Type) (Parameter 4)"
)
theta_history <- result$theta_history
iterations <- 1:nrow(theta_history)
# Plot the first parameter estimate
plot(iterations, theta_history[, 1], type = "l", col = "blue", lwd = 2,
     ylim = range(theta_history), ylab = "Parameter Estimate", xlab = "Iteration",
     main = "Parameter Estimates over Newton-Raphson Iterations")

# Add the lines for the remaining parameters
for (i in 2:ncol(theta_history)) {
  lines(iterations, theta_history[, i], col = i, lwd = 2)
}

# Add a legend to identify the parameters
legend("topright", legend = parameter_names,
       col = 1:ncol(theta_history), lwd = 2)
```
### Conclusion:

  - Convergence: The Newton-Raphson algorithm converges in about 4 iterations for all parameters, which is typical for well-behaved models.


## Comparson differences between NR and GLM estimates
```{r}
# Get final estimates from both methods
glm_estimates <- coef(modell)  # GLM estimates
nr_estimates <- result$final_theta  # NR estimates

#Calculate differences between NR and GLM estimates
diff_estimates <- abs(nr_estimates - glm_estimates)
# Print results for comparison
cat("GLM estimates:\n")
print(glm_estimates)

cat("NR estimates:\n")
print(nr_estimates)

cat("Differences between NR and GLM estimates:\n")
print(diff_estimates)
```
```{r}
#Plot the difference
library(ggplot2)
# Create a data frame for plotting
coef_names <- names(glm_estimates)
comparison_df <- data.frame(
  Method = rep(c("GLM", "NR"), each = length(glm_estimates)),
  Coefficient = rep(coef_names, 2),
  Estimate = c(glm_estimates, nr_estimates)
)
# Plot
ggplot(comparison_df, aes(x = Coefficient, y = Estimate, fill = Method)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Comparison of GLM and NR Estimates", y = "Coefficient Estimate", x = "Coefficient") +
  theme_minimal()

difference_df <- data.frame(
  Coefficient = coef_names,
  Difference = diff_estimates
)
```
```{r}
ggplot(difference_df, aes(x = Coefficient, y = Difference)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(title = "Absolute Differences between NR and GLM Estimates", 
       y = "Absolute Difference", x = "Coefficient") +
  theme_minimal()
```

# 3.Task3
```{r}
theta_hat <- result$final_theta

# Step 1: Calculate the Fisher Information matrix at the final theta estimates
fisher_info_hat <- I(theta_hat, y, X)

# Step 2: Invert the Fisher Information matrix
fisher_info_inv <- solve(fisher_info_hat)  # Inverse of the Fisher Information matrix

# Step 3: Extract the diagonal elements and compute the standard errors
std_errors_nr <- sqrt(diag(fisher_info_inv))  # Square root of the diagonal elements

# Print the NR algorithm's standard errors
cat("Standard errors from NR algorithm:\n")
print(std_errors_nr)

# Step 4: Compare with R's standard errors from glm()
# Get standard errors from R's glm model
std_errors_glm <- summary(modell)$coefficients[, "Std. Error"]

# Print the standard errors from glm()
cat("Standard errors from glm():\n")
print(std_errors_glm)

# Compare the two sets of standard errors
cat("Difference between NR and glm() standard errors:\n")
print(abs(std_errors_nr - std_errors_glm))

```
The output shows the comparison between the standard errors calculated using the Newton-Raphson (NR) algorithm and those provided by R’s `glm()` function. Here's what the results indicate:

### Observations:

- **Standard errors from NR algorithm** and **Standard errors from `glm()`** are nearly identical.
  - For each coefficient (`Intercept`, `Alder`, `KonMan`, and `UtbildareTrafikskola`), the standard errors computed by the NR algorithm match almost exactly with those from `glm()`.

- **Differences between NR and `glm()` standard errors** are extremely small, on the order of \(10^{-8}\) or \(10^{-9}\). This level of difference is negligible and likely due to minor numerical precision issues rather than any conceptual differences in the calculations.

### Conclusion:

- The **NR algorithm** and **R’s `glm()` function** produce nearly identical standard errors, which suggests that **R uses the same method** (i.e., the inverse of the Fisher Information matrix) to compute the standard errors for maximum likelihood estimates in logistic regression.
- The tiny differences (on the order of \(10^{-8}\)) are due to floating-point precision limitations inherent in numerical calculations, not due to any methodological differences.

Thus, it is safe to conclude that **R's `glm()` function computes standard errors in a similar way as your Newton-Raphson implementation**, confirming the equivalence of both methods in this context.

# 4.Task4
## comparison between the Bootstrap standard errors and the GLM standard errors
### Bootstrap standard errors
```{r}
# Number of bootstrap iterations
B <- 1000

```
```{r}
# Step 1: Fit the original model to get the initial coefficients and probabilities
model_orig <- glm(Resultat ~ Alder + Kon + Utbildare, 
                  data = data_individ, family = "binomial")

# Extract the original fitted probabilities
p_hat <- fitted(model_orig)

```
```{r}
# Extract the design matrix (X remains fixed during the bootstrap)
X_fixed <- model.matrix(Resultat ~ Alder + Kon + Utbildare, data = data_individ)

# Placeholder to store the bootstrap estimates
bootstrap_coefs <- matrix(NA, B, length(coef(model_orig)))
```
```{r}
# Define my own age and sex 
my_age <- 23  
my_sex <- "Kvinna"  
my_educ <- "Privatist"  
# Create a row for your specific case
X_individual <- c(1, my_age, ifelse(my_sex == "Male", 1, 0), ifelse(my_educ == "Privatist", 1, 0))

# Placeholder for bootstrap predictions for the individual case
bootstrap_predictions <- numeric(B)

```
```{r}
# Step 2: Run the bootstrap
set.seed(123)  # For reproducibility
for (b in 1:B) {
  # Resample the y-values based on the fitted probabilities
  y_boot <- rbinom(n = length(p_hat), size = 1, prob = p_hat)
  
  # Refit the model with the bootstrapped y-values
  model_boot <- glm(y_boot ~ Alder + Kon + Utbildare, family = "binomial", data = data_individ)
  
  # Store the estimated coefficients from the bootstrapped model
  bootstrap_coefs[b, ] <- coef(model_boot)
  
  # Calculate the predicted probability for the specific individual
  p_individual <- 1 / (1 + exp(-X_individual %*% coef(model_boot)))  # Logistic function
  bootstrap_predictions[b] <- p_individual
}

```
```{r}
# Step 3: Calculate the standard errors from the bootstrap
bootstrap_std_errors <- apply(bootstrap_coefs, 2, sd)

# Step 4: Compare the bootstrap standard errors with those from Task 3
cat("Bootstrap Standard Errors:\n")
print(bootstrap_std_errors)

```
```{r}
glm_std_errors <- summary(model_orig)$coefficients[, "Std. Error"]
std_errors_df_1 <- data.frame(
  Coefficient = c("Intercept", "Alder", "Kon", "Utbildare"),  # Adjust names accordingly
  GLM = glm_std_errors,
  Bootstrap = bootstrap_std_errors
)
print(std_errors_df_1)
```

```{r}
glm_std_errors <- summary(model_orig)$coefficients[, "Std. Error"]
# Combine the standard errors into a data frame for plotting
std_errors_df <- data.frame(
  Method = rep(c("GLM", "Bootstrap"), each = length(glm_std_errors)),
  Coefficient = rep(names(glm_std_errors), 2),
  StdError = c(glm_std_errors, bootstrap_std_errors)
)

ggplot(std_errors_df, aes(x = Coefficient, y = StdError, fill = Method)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Comparison of GLM and Bootstrap Standard Errors",
       y = "Standard Error", x = "Coefficient") +
  theme_minimal() +
  theme(legend.position = "top")
```

Conclusion:
- The two methods produce almost identical standard errors, indicating good stability and consistency in the model fitting.

- Reason for differences:
  - These differences arise because GLM calculates standard errors using the Fisher information matrix, while Bootstrap relies on repeated resampling, which is a non-parametric method. The small differences are statistically acceptable since Bootstrap provides a more flexible method that does not assume a specific distribution.
  - For example, the standard error for Alder in the GLM model is 0.008794838, while in Bootstrap, it's 0.008795902, showing only a minor difference. This indicates that the model parameters are fairly stable across both methods.

## comparison between the Bootstrap standard errors and the NR standard errors
```{r}
fisher_info_hat <- I(result$final_theta, y, X)  # Assuming I() is defined and computes the Fisher Information matrix
fisher_info_inv <- solve(fisher_info_hat)  # Inverse of the Fisher Information matrix
nr_std_errors <- sqrt(diag(fisher_info_inv))  # NR standard errors from the Fisher Information

std_errors_df_1 <- data.frame(
  Coefficient = c("Intercept", "Alder", "Kon", "Utbildare"),  # Adjust names accordingly
  NR = nr_std_errors,
  Bootstrap = bootstrap_std_errors
)

# Print the data frame to verify the result
print(std_errors_df_1)
```
```{r}
#Plotting
std_errors_df <- data.frame(
  Method = rep(c("NR", "Bootstrap"), each = length(nr_std_errors)),
  Coefficient = rep(c("Intercept", "Alder", "Kon", "Utbildare"), 2),  # Adjust names accordingly
  StdError = c(nr_std_errors, bootstrap_std_errors)
)
library(ggplot2)

ggplot(std_errors_df, aes(x = Coefficient, y = StdError, fill = Method)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Comparison of NR and Bootstrap Standard Errors",
       y = "Standard Error", x = "Coefficient") +
  theme_minimal() +
  theme(legend.position = "top")
```
Conclusion:
- The two methods produce almost identical standard errors, indicating good stability and consistency in the model fitting.

## Construct the 95% confidence interval for the individual probability
```{r}
# Step 5: Construct the 95% confidence interval for the individual probability
ci_lower <- quantile(bootstrap_predictions, 0.025)
ci_upper <- quantile(bootstrap_predictions, 0.975)
cat("95% Bootstrap Confidence Interval for the individual probability of success:\n")
cat("Lower bound:", ci_lower, "\n")
cat("Upper bound:", ci_upper, "\n")
```
- This confidence interval suggests that for an individual who is 23 years old, female, and privately educated, the probability of success (i.e., the response variable being 1) is between 51.07% and 65.03% with 95% confidence.
- This indicates a fairly high probability of success, with a narrow confidence interval, suggesting that the model’s prediction for this individual is relatively stable.
```{r}
# Plot the distribution of bootstrap predictions and highlight the confidence interval
hist(bootstrap_predictions, breaks = 30, main = "Bootstrap Predictions for Individual Probability of Success",
     xlab = "Predicted Probability", col = "lightblue", border = "white")

# Add vertical lines for the confidence interval bounds
abline(v = ci_lower, col = "red", lwd = 2, lty = 2)
abline(v = ci_upper, col = "red", lwd = 2, lty = 2)

# Add a label for the confidence interval on the plot
legend("topright", legend = c(paste("Lower CI (2.5%):", round(ci_lower, 4)),
                              paste("Upper CI (97.5%):", round(ci_upper, 4))),
       col = "red", lwd = 2, lty = 2, cex = 0.8)
```

