---
title: "Jiahui Li_computer assignment 2"
output:
  pdf_document:
    latex_engine: xelatex
date: "2024-10-05"
---

# Previous_Function
```{r}
source("funktioner.R")
#Likelihood function
L <- function(theta,y,X) {
  p <- 1/(1 + exp(-X %*% theta))  # logistic function
  likelihood <- prod(dbinom(y, size = 1, prob = p, log = FALSE)) # likelihood 
  return(likelihood)
}

#Log-likelihood function
l <- function(theta,y,X){
  p <- 1/(1+exp(-X %*% theta))
  log_likelihood <- sum((dbinom(y,size=1,prob=p,log= TRUE)))
  return(log_likelihood)
}

#Score function
S <- function(theta,y,X){
  p <- 1/(1+exp(-X %*% theta))
  score <- t(X) %*% (y-p)
  return(score)
}

#Fisher Information I function
I <- function(theta,y,X){
  p <- 1/(1+exp(-X %*% theta))
  v <- p * (1-p)
  D = diag(as.vector(v))
  fisher_info = t(X) %*% D %*% X
  return(fisher_info)
}

#Newton-Raphson
NR <- function(theta0, niter, y, X, tol = 0.001) {
  theta <- theta0  # Initialize parameters
  theta_history <- matrix(NA, niter + 1, length(theta0))  # Store parameter values for each iteration
  theta_history[1, ] <- theta0  # Store the initial value in history
  
  for (i in 1:niter) {
    score <- S(theta, y, X)  # Compute the score function
    fisher_info <- I(theta, y, X)  # Compute the Fisher Information matrix
    theta_new <- theta + solve(fisher_info) %*% score  # Update the parameters
    
    # Store the new parameter estimates
    theta_history[i + 1, ] <- theta_new
    
    # Compare the difference between new and old parameter values
    if (all(abs(theta_new - theta) < tol)) {
      cat("Converged after", i, "iterations\n")
      return(list(final_theta = theta_new, theta_history = theta_history[1:(i + 1), ]))
    }
    
    # Update theta with the new estimates and continue iterating
    theta <- theta_new
  }
  
  cat("Did not fully converge after", niter, "iterations\n")
  return(list(final_theta = theta, theta_history = theta_history))
}
```

# Data prepration
```{r}
load("proj_data.Rdata")
modell <- glm(Resultat ~ Alder + Kon + Utbildare, 
              data = data_individ,
              family = "binomial")
summary(modell)
```
```{r}
y <- matrix(data_individ$Resultat, ncol = 1)
X <- model.matrix(Resultat ~ Alder + Kon + Utbildare, 
                  data = data_individ)
```

# 1.Task1
```{r}
# Initial parameter guess
theta0 <- c(0,0,0,0) 

# Run the Newton-Raphson method to estimate parameters
theta_estimates <- NR(theta0, niter = 4, y = y, X = X) #actually niter=4 in the assignment1

# Extract the final estimated parameters
theta_estimates <- theta_estimates$final_theta
theta_estimates <- as.numeric(theta_estimates)

# Compute the Fisher Information matrix at the estimated parameters
fisher_info_estimates <- I(theta_estimates, y, X)

# Compute standard errors (square root of the diagonal of the inverse Fisher Information matrix)
standard_errors <- sqrt(diag(solve(fisher_info_estimates)))

# Compute Wald statistics (estimated parameters / standard errors)
wald_statistics <- theta_estimates / standard_errors
```
## Output the estimated parameters, standard errors, and Wald statistics
```{r}

theta_estimates  # Estimated parameters
```
(1) **Estimated Parameters (`theta_estimates`)**:
   - **Intercept**: 0.1547  
   - **Alder (Age)**: -0.0405  
   - **KonMan (Gender: Male)**: 0.3871  
   - **UtbildareTrafikskola (Driving School Education)**: 1.0987  
     This strongly positive coefficient suggests that having education from a driving school greatly increases the log-odds of the outcome.
```{r}
standard_errors  # Standard errors
```
(2) **Standard Errors (`standard_errors`)**:
     The standard error for education from a driving school is relatively small compared to the coefficient, indicating strong confidence in this estimate.

### comparison
```{r}

# Compare with the z-values from model summary
model_summary <- summary(modell)
coefficients <- model_summary$coefficients
z_values <- coefficients[, "z value"]

# Combine Wald statistics and z-values into a dataframe
comparison_df <- data.frame(Wald_Statistics = wald_statistics, Z_Values = z_values)
print(comparison_df)
```
### Summary
The results validate that the z-values in the glm output are indeed Wald statistics   calculated by dividing the estimated coefficients by their standard errors.
   - **Alder (Age)** has a highly significant negative impact on the outcome variable, as indicated by the large negative z-value.
   - **KonMan (Gender: Male)** shows a positive but moderate effect, with a z-value indicating significance but to a lesser extent compared to other variables.
   - **UtbildareTrafikskola (Driving School Education)** has the strongest positive effect on the outcome variable, as indicated by the high z-value, implying a very statistically significant impact.
   - The intercept shows a low z-value, suggesting it is not significantly different from zero and has no substantial impact.


# 2.Task2
```{r}
# Compute the log-likelihood of the full model
log_likelihood_full <- l(theta_estimates, y, X)

# Remove the gender variable and fit the reduced model
X_reduced <- X[, -3]  # Remove the 3rd column, which corresponds to "Kon" (gender variable)
theta0_reduced <- c(0, 0, 0)  # Initial parameter guess for the reduced model

#modify NR to complete iteration
NR1 <- function(theta0, niter, y, X, tol = 0.001) {
  theta <- theta0  # Initialize parameters
  theta_history <- matrix(NA, niter + 1, length(theta0))  # Store parameter values for each iteration
  theta_history[1, ] <- theta0  # Store the initial value in history
  
  for (i in 1:niter) {
    score <- S(theta, y, X)  # Compute the score function
    fisher_info <- I(theta, y, X)  # Compute the Fisher Information matrix
    theta_new <- theta + solve(fisher_info) %*% score  # Update the parameters
    
    # Store the new parameter estimates
    theta_history[i + 1, ] <- theta_new
    
    # Do not stop early based on tolerance
    theta <- theta_new  # Update theta with the new estimates and continue iterating
  }
  
  cat("Completed after", niter, "iterations\n")
  return(list(final_theta = theta, theta_history = theta_history))
}
eta_reduced <- NR1(theta0 = theta0_reduced, niter = 10, y = y, X = X_reduced)
log_likelihood_reduced <- l(eta_reduced$final_theta, y, X_reduced)

# Compute the generalized likelihood ratio statistic
likelihood_ratio_stat <- -2 * (log_likelihood_reduced - log_likelihood_full)

# Compare the squared Wald statistics with the generalized likelihood ratio statistic
wald_squared <- wald_statistics^2

# Chi-squared test with 1 degree of freedom (since only one parameter was removed)
df <- 1
p_value <- pchisq(likelihood_ratio_stat, df, lower.tail = FALSE)

# Output the results
wald_kon <- summary(modell)$coefficients["KonMan","z value"]
wald_kon_squared <- wald_kon^2
print("Wald Statistics Squared:")
print(wald_squared)

print("Wald Statistics Kon Squared:")
print(wald_kon_squared)

print("Likelihood Ratio Statistic:")
print(likelihood_ratio_stat)

print("P-value for Likelihood Ratio Test:")
print(p_value)

```
## Conclusion:
  - The Wald Statistics Kon Squared and Likelihood Ratio Statistic are approximately equal.
  - The "P-value for Likelihood Ratio Test" is 0.004839648. This p-value indicates that at the 5% significance level, the result of the likelihood ratio test is significant, meaning that the null hypothesis (that the coefficient of Kon is equal to 0) can be rejected. In other words, the variable Kon is significant in the model.
  
# 3.Task3
```{r}
# Define the score function
S_null <- function(theta, y, X) {
  p <- 1 / (1 + exp(-X %*% theta))  # Compute logistic regression probabilities
  score <- t(X) %*% (y - p)  # Compute the score function
  return(score)
}

# Define the Fisher Information matrix function
I_null <- function(theta, y, X) {
  p <- 1 / (1 + exp(-X %*% theta))  # Logistic regression probabilities
  v <- p * (1 - p)  # Variance term
  D <- diag(as.vector(v))  # Diagonal matrix
  fisher_info <- t(X) %*% D %*% X  # Compute the Fisher Information matrix
  return(fisher_info)
}

# Get y and X from the dataset
y <- matrix(data_individ$Resultat, ncol = 1)

# Construct the design matrix, including all variables (Alder, Kon, Utbildare)
X_full <- model.matrix(Resultat ~ Alder + Kon + Utbildare, data = data_individ)

# Assume intercept and gender (Kon) are 0, i.e., use as an offset
offset_term <- rep(0, nrow(data_individ))  # Offset is set to zero

# Fit the model using the glm() function, only estimating Alder and Utbildare
reduced_model <- glm(Resultat ~ Alder + Utbildare, 
                     data = data_individ, 
                     family = "binomial", 
                     offset = offset_term)

# Extract the maximum likelihood estimates of the nuisance parameters
eta_ml <- coef(reduced_model)

# Create the design matrix including only Alder and Utbildare
X_reduced <- model.matrix(~ Alder + Utbildare, data = data_individ)

# Compute the score function under the null hypothesis
score_null <- S_null(eta_ml, y, X_reduced)

# Compute the Fisher Information matrix under the null hypothesis
fisher_info_null <- I_null(eta_ml, y, X_reduced)

# Compute the generalized score statistic TS(θ0)
score_statistic <- t(score_null) %*% solve(fisher_info_null) %*% score_null
score_statistic <- as.numeric(score_statistic)  # Convert to numeric type

# Degrees of freedom for the chi-squared distribution is 2 (since two parameters are constrained: intercept and Kon)
df <- 2

# Compute the p-value based on the chi-squared distribution
p_value_score_statistic <- pchisq(score_statistic, df, lower.tail = FALSE)

# Output the generalized score statistic and p-value
print("Generalized Score Statistic:")
print(score_statistic)

print("P-value for Generalized Score Test:")
print(p_value_score_statistic)


```
### Summary of the Results:

1. **Generalized Score Statistic (`score_statistic`)**:
   - The value of the score statistic is extremely small, approximately \(2.91667 \times 10^{-22}\).
   - This near-zero value indicates that the model fits very well under the null hypothesis. In other words, **removing the intercept and gender (Kon) variables does not significantly affect the model's fit**.

2. **p-value (`p_value_score_statistic`)**:
   - The calculated p-value is **1**, which means the null hypothesis is not rejected.
   - In statistical testing, a p-value of 1 implies **no significant reason to reject the null hypothesis**. This means that removing the intercept and gender variables does not lead to any significant degradation in the model's performance.

## Conclusion:

- The combination of a **near-zero generalized score statistic** and a **p-value of 1** indicates that **removing the intercept and gender (Kon) variables does not significantly impact the model’s fit**. The model fits well under the null hypothesis, meaning these two variables have little to no influence on the outcome.
  
- **Accepting the null hypothesis** suggests that the intercept and gender variables can be set to zero without worsening the model's performance. In other words, these variables **do not make a significant contribution to the predictive power of the model**.

# 4.Task4
```{r}
# Define a grid of theta_Kon values
theta_Kon_grid <- seq(-2, 2, length.out=100)  
log_likelihoods <- numeric(length(theta_Kon_grid))
profile_likelihoods <- numeric(length(theta_Kon_grid))

# list every theta_Kon
for (i in seq_along(theta_Kon_grid)) {
  theta_kon_present <- theta_Kon_grid[i]
  offset_term <- X[, 3] * theta_kon_present
  model <- glm.fit(x = X[, -3], y = y,
                       family = binomial(),
                       offset = offset_term)
  #log-likelihood for the fitted model
  log_likelihoods[i] <- sum(dbinom(y, size = 1,
                                   prob = model$fitted.values,
                                   log = TRUE))
  # Profile likelihood
  profile_likelihoods[i] <- exp(log_likelihoods[i])
}
# Plot the profile likelihood
plot(theta_Kon_grid, profile_likelihoods, type = "l", col = "blue",
     main = "Profile Likelihood for theta_Kon",
     xlab = expression(theta[Kon]),
     ylab = "Profile Likelihood")

# Add a horizontal line for 95% confidence interval level
max_likelihood <- max(profile_likelihoods)
ci_level <- max_likelihood * exp(-0.5) # exp(-0.5) is the drop for 95% CI
abline(h = ci_level, col = "red", lty = 2)
legend("topright", legend = c("Profile Likelihood", "95% CI Level"),
       col = c("blue", "red"), lty = 1:2)

coef_estimates <- summary(modell)$coefficients

# Wald intervals for theta_Kon
theta_kon_estimate <- coef_estimates["KonMan", "Estimate"]
theta_kon_se <- coef_estimates["KonMan", "Std. Error"]
wald_left <- theta_kon_estimate - 1.96 * theta_kon_se # 95% confidence interval
wald_right <- theta_kon_estimate + 1.96 * theta_kon_se

# Add Wald interval lines 
abline(v = wald_left, col = "green", lty = 2)
abline(v = wald_right, col = "green", lty = 2)
legend("topright", legend = c("Profile Likelihood", "95% CI Level", "Wald CI"),
col = c("blue", "red", "green"), lty = 1:2)
```
```{r}
# Print the Wald interval
cat("Wald Interval for theta_Kon: [", wald_left, ", ", wald_right, "]\n")
```
## Conclusion:
- The Wald test provides a confidence interval of `[0.1161208, 0.658176]`.
- The **profile likelihood interval** is more concentrated, suggesting that it might be a more accurate estimation than the Wald interval.
- This indicates that the estimated coefficient for the gender variable \(\theta_{\text{Kon}}\) is likely to fall between 0.2 and 0.6 with 95% confidence.
